White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-01-21 papers/wdmerger_II/paper.tex
Add a description to paper II of the new burning limiter

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here.
(A related approach presented in \citet{kushnir:2013} is to limit the magnitude
of the energy release in any given integration by scaling all the reaction rates.)
At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum burning
timestep. In a given zone, the burning timestep is defined as:
\begin{equation}
  \Delta t_{\text{burning}} = f_b\, \frac{e}{\dot{e}}.
\end{equation}
Here $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning. We estimate this using the energy change in the last burning
update, divided by the length of the burning timestep. The factor $f_b$ then
determines by what fraction we would like to allow the internal energy to change
in the current timestep (assuming that $\dot{e}$ does not change quickly from
timestep to timestep, which is not always the case). By making $f_b$ smaller,
we can control the magnitude of the decoupling between the reactions and the hydro.
A typical choice in the literature is $f_b \sim 0.3$. This timestep limiting has
now been implemented in \castro\ and the factor $f_b$ can be set at runtime.

>>>>

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here.
We define a burning-limited timestep $\Delta t_{b1}$ as:
\begin{equation}
  \Delta t_{b1} = f_{b1}\, \frac{e}{|\dot{e}|}.
\end{equation}
Here $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning. We estimate this using the energy change in the last burning
update, divided by the length of the burning timestep. The factor $f_{b1}$ then
determines by what fraction we would like to allow the internal energy to change
in the current timestep (assuming that $\dot{e}$ does not change quickly from
timestep to timestep, which is not always the case). By making $f_{b1}$ smaller,
we can control the magnitude of the decoupling between the reactions and the hydro.
Typical choices in the literature are $f_{b1} = 0.2$ or $0.3$. This timestep limiting has
now been implemented in \castro\ and the factor $f_{b1}$ can be set at runtime.

\citet{kushnir:2013} point out this is not the only way for the burning to
become numerically unstable; another failure mode is when the energy injection timescale
is faster than the sound-crossing time in a zone. \citet{kushnir:2013} solved this by
limiting the magnitude of the energy release after a burn; we take a slightly different
approach and limit the timestep based on the previous burning, as with the other limiter.
The goal is to ensure that the following condition holds:
\begin{equation}
  \frac{\partial e}{\partial t} = f_{b2} \frac{e}{t_s},
\end{equation}
The sound crossing time, $t_s$, is given by $\Delta x / c_s$, where $c_s$ is the sound speed
and $\Delta x$ is the zone size. The parameter $f_{b2}$ then determines the ratio of the
nuclear energy generation timescale to the sound-crossing time, in the limit of uniform
timesteps. \citet{kushnir:2013} choose $f_{b2} = 0.1$ for their simulations.
(As with the previous limiter, this parameter can be set at runtime in \castro.)
We then re-write this as a limit on the size of the next timestep, using the same
estimate as above for the nuclear energy generation rate $\dot{e}$ from the last burn:
\begin{equation}
  \Delta t_{b2} = \left[f_{b2}\, \frac{e}{|\dot{e}| \Delta t_{\rm{old}}}\right]^{-1}\, {t_s}
\end{equation}
Here $\Delta t_{\rm{old}}$ is the size of the timestep used in the burning advance that calculated
$\dot{e}$. In the case of our Strang splitting approach it is half of the old advance
timestep, $\Delta t^{n-1} / 2$. Thus the denominator inside the brackets is simply the net
change in internal energy during the last burn.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of each
burning timestep across all zones.
